Nature or Nurture? Collisionless Evolution of Galactic Disc-Halo Systems

### 2019-07-15 Thesis/Conclusion.tex
Writing for July 12-14

We found that modelling the disk in a dynamic fashion is tremendously important for the inner halo. Specifically, we found the formation of a bar led to a more concentrated halo mass distribution.

@@

@@

@@

\subsection{Future Work}
\section{Vetical Structure in Cosmological Stellar Disks}

>>>>

We find that modelling the disk in a dynamic fashion is tremendously important for the inner halo. Specifically, we found the formation of a bar led to a more concentrated halo mass distribution. On the whole, these results suggest that if one is interested in the Core-Cusp problem, this should be studied with fully live models of stellar disks.

@@

This last aspect of our work has informed some design decisions by others, namely in the work of \citet{read_2019}. If one is just interested in subhalo population statistics, it is sufficient to model baryons as fixed potentials in a pure dark matter halo.

Beyond the effect of the disk on the halo, we find that our rigid disk approximation is quite realistic in modelling the angular evolution of the disk. This allows us to set up near-equilibrium cosmological systems without the use of MHD simulations. Since the work in this thesis focuses on the underlying theoretical mechanisms driving the evolution of galaxies in cosmological environments, we only need a good representation of the true systems.

@@

Our algorithm makes some optimizations over previous disk insertion schemes. The optimizations we made are by no means exhaustive, and a detailed discussion of potential optimizations is given in Chapter~\ref{ch:implementation}. We should consider a scheme which extracts mass from the halo particles during the disk growth phase. Dark matter particles in a dark matter only simulation include the baryonic mass of the Universe. Since the disk and halo contribute on the same order to the potential in the inner parts of a galaxy, it would be more realistic to subtract mass from halo particles as the disk grows. This may have a sizable impact on our conclusions about the disk's effect on adiabatic contraction.

Finally, we might also consider a scheme which initializes the disk with some gas component. \citet{deg_2019} modified \textsc{GalactICS} to allow for the inclusion of gas. This, combined with modifying halo particle masses, should allow us to more accurately compare our results to fully hydrodynamical \textit{ab initio} simulations.

Even with the current algorithm, there is more we can do on the study of disk-halo interactions. A possible project would be to study the effect of varying disk properties in the same halo. We might also consider adding a spheroidal bulge component to see how high baryonic mass concentrations change our results. We might also look for more detailed evidence of the halo's response to the disk in our existing simulations.

@@

We discuss our contributions to understanding how bars form in stellar disks in Chapter~\ref{ch:paper_ii}. The paper unifies previous work on disk insertion schemes by viewing them through a lens of a common parameter space. In doing so, we illustrate that various discrepancies between our work and previous work are actually due to selection effects in this space.

In particular, we show that a commonly discounted variable, the disk thickness, is a primary driver of bar strength in a cosmological setting. We show that, in general, the length of a bar at present day for cosmological stellar disks is largely a function of the initial disk thickness. This effect is not numerical, as we show through our application of the \textsc{AGAMA} code.

The importance of being cautious about the choice of gravitational softening length is also illustrated. Since disk thickness appears to play a major role in bar formation in $\Lambda$CDM, we must be careful that we are resolving our disks' vertical structure.

On the whole, we show that it is difficult to suppress the formation of bars in a cosmological setting. While the bar more closely resemble the Milky Way's when we start with a thinner disk, we find strong bars form nonetheless. As such, there is still a discrepancy in accounting for the observed bar fraction without invoking classical bulges or non-collisionless phenomena. Our work has proved useful in illustrating this, and we are credited for furthering discussion of this issue in \citet{sellwood_2019}.

A key flaw of the work in Chapter~\ref{ch:paper_ii} is that we only considered a single halo. In Chapter~\ref{ch:paper_iii}, we ran twelve simulations in four halos. Although all of the halos had the same total mass as the Milky Way, they varied widely in terms of their other properties. Despite the differences in the halos, the bars formed by embedded disks were quite similar. Additionally, there is a nontrivial dependence on Toomre $Q$ in terms of when the bar buckles, but the final bar strength itself does not depend on this parameter.

In Chapter~\ref{ch:paper_iii}, bar formation was not the focus of the study. Now that we have a sample of bars forming in different cosmological environments, we can position the disks in the parameter space described in Chapter~\ref{ch:paper_ii}. A proper synthesis with previous literature on this topic should be performed.

An interesting feature of our simulations was the existence of ``banana bars" in some of the simulations. Understanding why we get banana bars in some simulations versus a more traditional X-shaped pattern might provide an observational constraint for the Milky Way. This topic should be explored in more detail.

A valid criticism of our disk insertion technique is that all of the stars are initialized simultaneously. This is unrealistic, and allowing stars to gradually form out of the rigid disk potential might show a more complicated dependence on the initial conditions parameter space. We think developments in this direction would be more directly comparable to \textit{ab initio} simulations, and present a vector for future research.

\section{Vetical Structure in Cosmological Stellar Disks}
\subsection{Summary}

The main focus of Chapter~\ref{ch:paper_iii} was how disks form vertical structure in $\Lambda$CDM environments. We showed that there are a wide variety of Milky Way-mass halos that form vertical structure consistent with structure observed in the solar neighborhood. In particular, we showed that this structure can be formed without a massive satellite encounter in a cosmological setting. This position is bolstered by the fact that we see Monoceros-like structure in all twelve of our simulations.

We also found that consistent perpendicular torques on the disk are required to explain features in the outer disk. It is unlikely that substructures far below the $10^{11}$ solar mass regime are responsible for stars stripped to higher latitudes. Global tides from disk-halo axis misalignment seem to be the most effective way to create these populations of stars. We therefore suggest that streams originating from the outer disk may primarily depend on the Milky Way's interaction with the LMC. A massive merger like the one with the LMC can cause consistent perpendicular torques its own tides or by disturbing the existing smooth Milky Way halo mass.

A key flaw in this work was that we had no LMC analogues. An interesting experiment would be to compare the vertical structure caused by a future LMC merger and the vertical structure generated when a disk is misaligned in a triaxial halo. On the whole, future work should proceed in the same way as the work in Chapter~\ref{ch:paper_ii}. We should seek to relate the halo environments to a common parameter space.

We also might want to compare our simulations in the solar neighborhood to Gaia data through simulation-generated mock catalogues. Although not presented in this thesis, we have developed a code which allows us to generate initial conditions that are selectively upsampled in a specified annulus. In isolated galaxy simulations, the upsampled annulus remains mostly uncontaminated at late times. By resampling the solar neighborhood, and by using simulation-generated mock catalogues, we can compare our simulations more directly with observations.

\section{Closing Remarks}

\bibliographystyle{apalike}
\bibliography{bibliography_conclusions.bib}

### 2019-07-15 Thesis/Thesis.tex
Writing before lunch on June 27
Writing for July 12-14

%*************************************************************************************************************
% ABSTRACT
%*************************************************************************************************************

@@

@@

Blah blah blah.

Only required by CHEM, COMPUTING, GEOL, MATH and Physics (Ph.D. ONLY!).

@@

%*************************************************************************************************************
% MEATY CHAPTERS
%*************************************************************************************************************

@@

% Include my chapter texts - kept separated to make editing easier.
%\include{Glossary}
\include{Introduction}
\include{Background}
\include{Paper_I}
\include{Paper_II}
%\include{Alloy}

>>>>

\begin{center}
\vspace*{\stretch{1}}
\noindent To Mom and Dad. Thank you for supporting me when I have been hard to support.

-- Your Loving Son
\vspace*{\stretch{1}}
\end{center}

\clearpage

%{
%\noindent }%*************************************************************************************************************
% ABSTRACT
%*************************************************************************************************************

@@

The research presented in this thesis was completed under the supervision of Prof.
Lawrence M. Widrow at Queen’s University. All of the work presented here was done
by the author (Jacob S. Bauer) except where explicitly stated otherwise.

Chapter~\ref{ch:paper_i} contains a reproduction of a paper published in  Monthly Notices of the
Royal Astronomical Society as: Jacob S. Bauer, Lawrence M. Widrow, and Denis Erkal. Disc-halo interactions in $\Lambda$CDM. \textit{Monthly Notices of the Royal Astronomical Society}, 476:198-209,
2018. For this paper, I ran all of the simulations. The writing is mine with direct input from Lawrence M. Widrow and Denis Erkal. I performed all of the calculations. The analysis was handled primarily by me under the direction of Lawrence M. Widrow, with substantial input from Denis Erkal.

Chapter~\ref{ch:paper_ii} contains a reproduction of a paper published in  Monthly Notices of the
Royal Astronomical Society as: Jacob S. Bauer and Lawrence M. Widrow. Can stellar discs in a cosmological setting avoid forming strong bars?. \textit{Monthly Notices of the Royal Astronomical Society}, 476:523-537,
2019. For this paper, I ran all of the simulations. The writing is mine with direct input from Lawrence M. Widrow. I performed all of the calculations, except the one in \S2.3 which was performed by Lawrence M. Widrow. Figures 2-4 were also the work of Lawrence M. Widrow. The analysis was handled primarily by me under the direction of Lawrence M. Widrow.

Chapter~\ref{ch:paper_iii} contains a draft of a paper in preparation to be submitted to Monthly Notices of the Royal Astronomical Society. The working title for the paper is ``Vertical Structure and Kicked-Up Stars in $\Lambda$CDM Galaxies". For this work, I performed all of the simulations and writing. The analysis was handled primarily by me under the direction of Lawrence M. Widrow.

@@

As I reach the end of this chapter of my life, I know that I did not reach this milestone unassisted. Here, I recognize those who have helped me in this accomplishment.

I want to acknowledge, first and foremost, the amazing supervision I received from Larry Widrow. His careful, unyielding approach to scientific research has not only shown me how to be a better scientist, but a more conscientious communicator. I am immensely grateful for the opportunity to work with him, and for the years of guidance that will serve me well for the rest of my life.

In addition to Larry, I want to acknowledge my other mentors, Heidi Newberg and Denis Erkal. Through my work with each of them, I learned how to evaluate myself more honestly, both professionally and personally. I am grateful that through my work with my mentors, I learned to be simultaneously self-critical and able to forgive myself for mistakes. Without Heidi's and Denis' guidance, I would not have been able to finish this work.

To all of the other faculty members and Department staff, I am tremendously grateful for your contributions to me personally, and to the program as a whole. I am proud of my friends and colleagues at Queen's, and it is through your hard work that this positive academic environment exists. I wish you all the best.

I am grateful for the years of friendship with my fellow astronomy graduate students. In their own ways, they have all kept me mostly sane for 4 years. I would not have been able to perservere without their support. Although they did not succeed in turning me into a Canadian, they did manage to make me grow a beard. I look forward to many more years of friendship.

I want to acknowledge my wonderful family for their support. Mom and Dad have been my biggest supporters through the ups and downs, and this thesis is as much their accomplishment as it is mine.

I am grateful to Rachel Peckman, Chris Weir, and Taylor Weir for our lasting friendships. In high school, Chris showed me that I was capable of pursuing science. Our adventures in Tech Club (RIP), our shenanigans at RPI (RIP), and subsequent friendship (not RIP) made me strive to be a better technologist and person.

Lastly, I want to acknowledge my partner, Alexis Hill. I can't wait to see what adventures life hold for us in Austin. I wouldn't want to be on this roller coaster called life with anyone else.

@@

\chapter*{List of Abreviations}
\begin{longtable}{ll} 
\hline
Abbreviation & Expansion \\ \hline
\textsc{AGAMA} & Action-based Galaxy Modelling Architecture \\
DF & Distribution Function \\
F(L)RW & Friedmann (Lema\^itre) Robertson Walker\\
LAMOST & Large Sky Area Multi-Object Fiber Spectroscopic Telescope \\
M31 & Messier object 31 (Andromeda) \\
M33 & Messier object 33\\
NFW & Navarro-Frenk-White\\
RAVE & Radial Velocity Experiment \\
SDSS & Sloan Digital Sky Survey\\
Sgr dSph & Sagittarius Dwarf Spheroidal [Galaxy]\\
Sgr  & [The] Sagittarius [Stream]\\ \hline
\end{longtable}

\chapter*{List of Symbols}
\begin{longtable}{ll} 
\hline
Symbol/Unit & Meaning\\ \hline
Gyr & Gigayear; $10^9$ years\\
kg  & Kilogram; standard SI unit of mass\\
kpc & Kiloparsec; $10^3$ parsecs\\
m   & meter; standard SI unit of length\\
$M_\odot$ & Solar mass; approximately $2 \times 10^{30}$ kilograms\\
Mpc & Megaparsec; $10^6$ parsecs\\
Myr & Megayear; $10^6$ years\\
$O(f)$ & Big O notation; on the order of $f$ \\
pc  & parsec; approximately $3.1 \times 10^{16}$ meters\\
s   & second; standard SI unit of time\\
yr  & year; approximately $3.2 \times 10^7$ seconds\\ \hline
\end{longtable}
%*************************************************************************************************************
% MEATY CHAPTERS
%*************************************************************************************************************

@@

% Include my chapter texts - kept separated to make editing easier.
%\include{Glossary}
\include{Introduction}
\include{Background}
\include{Paper_I}
\include{Implementation}
\include{Paper_II}
\include{Paper_III}

%\include{Alloy}
